Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (*Sida fallax*)

### 2023-10-28 ms/ms.qmd
revising and added habitat-climate analysis

@@

```{=tex}
\begin{figure}[H]
  \includegraphics{../figures/study-system.pdf}
  \caption{(A) Typical growth form of montane (left) and coastal (right) ‘ilima plants and (B) leaves. (C) Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island).}
  \label{fig:study-system}
\end{figure}
```

@@

## Plant sampling --- {#plant-sampling}

We identified 7 suitable natural populations of ʻilima on Oʻahu and 5 on Hawaiʻi Island by consulting @yorkston_interfertility_2006 and citizen scientist records on iNaturalist [@noauthor_inaturalist_2022] (\autoref{fig:study-system}C; \autoref{tbl-locations}). We avoided sites that appeared to be cultivated. We visited sites between August and November 2022. For logistical reasons, the sites on Hawaiʻi were sampled during one three-day trip. We haphazardly sampled eight plants distributed evenly between the highest and lowest elevation plants along a transect at each site. For safety and conservation reasons, transects were along a trail or road. We did not sample small individuals if there was risk removing leaves would cause mortality. From each plant, we collected two fully expanded leaves for trait measurements. We sampled stomatal traits on all leaves; leaf thickness on one leaf from three randomly selected plants per site; and, due to limited time, a single leaf from a single plant at the middle of each transect for gas exchange measurements.

@@

The log-response ratio is commonly used social and biological sciences (e.g. @hedges_metaanalysis_1999). It is straightforward to interpret because values above 0 indicate a photosynthetic advantage of amphistomy, whereas values less than 0 indicate a disadvantage. The response ratio is also better than the absolute difference because it indicates a proportional change in $A$, which facilitates comparisons across leaves and environments with different baseline photosynthetic rates. The irradiance of the light source in the pseudohypo leaf is higher because the propafilm reduces transmission. To compensate for reduced transmission, we increased incident $\mathrm{PPFD}$ for pseudohypo leaves by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm. We also set the stomatal conductance ratio, for purposes of calculating boundary layer conductance, to 0 for pseudohypo leaves following manufacturer directions.

To control for order effects, we alternated between starting with amphi (untreated) or pseudohypo (treated) leaf measurements. In the field, we acclimated the focal leaf to high light ($\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), high relative humidity ($\mathrm{RH} = 70\%$), $T_\text{leaf} = 25-30~° \text{C}$, depending on ambient temperature, until $A$ and $g_\text{sw}$ reach their maximum. We used these data as our estimates of maximum $A$ and $g_\text{sw}$. After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without photosynthetic downregulation. Hence, $A_\text{amphi}$ and $A_\text{hypo}$ were both measured at low chamber humidity after the leaf had acclimated to high humidity. All other environmental conditions in the leaf chamber remained the same. We logged data until $g_\text{sw}$ reached its nadir. We then repeated the process with the other treatment (amphi or pseudohypo).

@@

@@

```{=tex}
\begin{figure}[H]
  \includegraphics{../figures/habitat-aa.pdf}
  \caption{Coastal leaves benefit more amphistomy than montane leaves. A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$. Each point-interval is the median posterior estimate plus 95\% confidence interval of amphistomy advantage for that leaf. Each leaf is from a different montane (blue) or coastal (orange) site, arranged by habitat and ascending amphistomy advantage within habitat. The longer horizonal bars are the average amphistomy advantage for montane and coastal leaves. $g_\mathrm{sw}$, stomatal conductance to water vapor.}
  \label{fig:habitat-aa}
\end{figure}
```
## Leaf thickness is associated with amphistomy advantage between but not within habitats ---

@@

Amphistomy is a seemingly simple way that leaves can increase carbon gain without significant additional water loss, yet it is rare in nature and we do not know why. The strong association between amphiostomy and sunny, open habitats suggests that amphistomy may benefit sun leaves more than shade leaves, but progress has been limited by the lack of evidence that amphiostomy *per se* affects improves photosynthesis in a given leaf. By experimentally blocking gas exchange through the upper surface in a controlled environment, we directly compared an amphistomatous leaf to an otherwise identical pseudohypostomatous leaf. This allows us to quantity the amphistomy advantage ($\mathrm{AA}$) holding all else constant. Taking advantage of the steep climatic gradients in the Hawaiian archipelago, we applied this new method to show for the first time that sun leaves benefit  `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy than shade leaves on ʻilima (*Sida fallax*) plants ($\mathrm{AA}_\text{coastal} = `r prettyNum(habitat_aa[['coastal']], digits = 2L)`$ vs. $\mathrm{AA}_\text{montane} = `r prettyNum(habitat_aa[['montane']], digits = 2L)`$). While this result has not yet been validated in other species, our results indicate that part of the reason amphistomatous leaves are found most commonly in high light habitats is that the adaptive benefit is greater in this environment.

If $\mathrm{AA}$ is typically greater in sun leaves than shade leaves, it could partially explain the distribution of amphi- and hypostoamtous leaves, but the precise mechanism(s) require further study. One hypothesis is that the internal airspace conductance, $g_\mathrm{ias}$, from stomata to mesophyll cell walls is lower in thicker sun leaves [@parkhurst_adaptive_1978]. All else being equal, a leaf with lower $g_\mathrm{ias}$ will benefit more from amphistomy. Our results partially support this hypothesis. Coastal ʻilima leaves with high $\mathrm{AA}$ (\autoref{fig:habitat-aa}) are thicker than montane leaves, but the relationship between $\mathrm{AA}$ and leaf thickness within habitats is actually slightly negative (\autoref{fig:traits-aa}A), opposite our prediction. Since coastal and montane leaves differ in many respects besides thickness, we do not have enough data to conclude that leaf thickness explains the variation in $\mathrm{AA}$ between habitats. Alternatively, other biochemical or anatomical differences between coastal and montane leaves may explain why $\mathrm{AA}$ is greater in coastal leaves. The negative relationship, (albeit nonsignificant in that 95\% confidence intervals encompassed 0), between leaf thickness and $\mathrm{AA}$ could be explained if thicker leaves compensated by having a more porous mesophyll and/or less tortuous airspaces [@theroux-rancourt_maximum_2021].

A second natural hypothesis is that amphistomatous leaves with few adaxial (upper) stomata benefit less than those with similar densities on both surfaces. We predicted that leaves with $g_\mathrm{smax,ratio}$ closer to 0.5 would have higher $\mathrm{AA}$ based on biophysical models [@gutschick_photosynthesis_1984]. The logic is that a small number of stomata on the upper surface are insufficient to supply the entire upper mesophyll due to limited lateral diffusion [@morison_lateral_2005]. Our results do not support this hypothesis. Montane leaves from Big Island sites had low $g_\mathrm{smax,ratio}$ and low $\mathrm{AA}$ whereas low montane leaves on Oʻahu had high $g_\mathrm{smax,ratio}$, but similarly low $\mathrm{AA}$ (\autoref{fig:traits-aa}B). Among coastal sites, the site with the lowest $g_\mathrm{smax,ratio}$ had the lowest $\mathrm{AA}$, but there was little variation in $g_\mathrm{smax,ratio}$ among coastal leaves in our sample. We therefore cannot rule out that a larger sample of coastal leaves with greater variance in $g_\mathrm{smax,ratio}$ might support this hypothesis.

@@

Two limitations of this study are the small sample size and experimental design that precludes distinguishing genetic from environmental differences in leaf traits. Understanding the mechanistic basis of higher $\mathrm{AA}$ in sun leaves would require much larger sample sizes. Sun leaves tend to be thicker, more densely packed with mesophyll cells, and have greater photosynthetic capacity and higher stomatal conductance, among other traits [@lambers_plant_2008]. Each of these factors and others potentially modulate $\mathrm{AA}$. Quantifying the contribution of all these factors requires larger samples and additional measurements that are beyond the scope of this study, but exciting avenues for future research on leaf structure-function relations. Although many morphological traits that distinguish coastal and montane ʻilima populations persist in a common environment [@yorkston_interfertility_2006], we cannot distinguish between genetic effects and plastic responses to habitat as causes of difference in $\mathrm{AA}$ because we measured naturally occurring plants *in situ*. While disentangling genetic and plastic contributions is not necessarily important for understanding the distribution of amphistomatous leaves, it would be insightful to know about genetic and environmental contributions to trait variation. A reciprocal transplant would be able to determine the genetic and environmental contributions, as well their interaction, to trait variance in nature.

>>>>

# From 16_fit-habitat-climate.R
df_stat = read_rds("../objects/habitat-climate-stats.rds")

@@

```{=tex}

\begin{figure}[H]
  \includegraphics[height=3.25in]{../figures/study-system.pdf}
  \includegraphics[height=3.25in]{../figures/habitat-climate.pdf}
  \caption{A. Typical growth form of montane (left) and coastal (right) ‘ilima plants and B. leaves. C. Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island). D. Climatic, light, and vegetation height comparisons between montane (blue) and coastal (orange) habitats sampled in this study. Open circles are values for the midpoint of each site transect; closed circles and intervals are the mean $\pm~1$ standard error. The habitats did not differ significantly in mean annual precipitation (top-left) or solar radiation (bottom-left), but montane plants inhabit significantly cooler climates (top-right) with greater vegetation height (bottom-right). MAP = mean annual precipitation; MAT = mean annual temperature; ns = not significant; * indicates $0.01 \le P < 0.05$; ** indicates $0.001 \le P < 0.01$.}
  \label{fig:study-system}
\end{figure}

@@

## Plant sampling and climate --- {#plant-sampling}

We identified 7 suitable natural populations of ʻilima on Oʻahu and 5 on Hawaiʻi Island by consulting @yorkston_interfertility_2006 and citizen scientist records on iNaturalist [@noauthor_inaturalist_2022] (\autoref{fig:study-system}c; \autoref{tbl-locations}). We avoided sites that appeared to be cultivated. We visited sites between August and November 2022. For logistical reasons, the sites on Hawaiʻi were sampled during one three-day trip. We haphazardly sampled eight plants distributed evenly between the highest and lowest elevation plants along a transect at each site. For safety and conservation reasons, transects were along a trail or road. We did not sample small individuals if there was risk removing leaves would cause mortality. From each plant, we collected two fully expanded leaves for trait measurements. We sampled stomatal traits on all leaves; leaf thickness on one leaf from three randomly selected plants per site; and, due to limited time, a single leaf from a single plant at the middle of each transect for gas exchange measurements. We downloaded climatic data on mean annual temperature, solar radiation, and vegetation height from the Climate and Solar Radiation of Hawaiʻi databases [@giambelluca_evapotranspiration_2014] using the latitude and longitude at the middle of each transect. We also downloaded mean annual precipitation from 1978-2007 from the Rainfall Atlas of Hawaiʻi [@giambelluca_online_2013]. The spatial resolution is approximately $234 \times 250$m. The temperature data are calibrated from networks of meteorological stations operating in the late 20th century and 21st century; the solar radiation data are calibrated from satellite measurements collected between 2002 and 2009 [@giambelluca_evapotranspiration_2014]. We tested whether climatic variables differed among our coastal and montane populations using Welch's two-sample $t$-test.

@@

The log-response ratio is commonly used social and biological sciences (e.g. @hedges_metaanalysis_1999). It is straightforward to interpret because values above 0 indicate a photosynthetic advantage of amphistomy, whereas values less than 0 indicate a disadvantage. The response ratio is also better than the absolute difference because it indicates a proportional change in $A$, which facilitates comparisons across leaves and environments with different baseline photosynthetic rates. The irradiance of the light source in the pseudohypo leaf was higher because the propafilm reduces transmission. To compensate for reduced transmission, we increased incident $\mathrm{PPFD}$ for pseudohypo leaves by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm. We also set the stomatal conductance ratio, for purposes of calculating boundary layer conductance, to 0 for pseudohypo leaves following manufacturer directions.

To control for order effects, we alternated between starting with amphi (untreated) or pseudohypo (treated) leaf measurements. In the field, we acclimated the focal leaf to high light ($\mathrm{PPFD} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}$), high relative humidity ($\mathrm{RH} = 70\%$), $T_\text{leaf} = 25-30~° \text{C}$, depending on ambient temperature, until $A$ and $g_\text{sw}$ reach their maximum. We used these data as our estimates of maximum $A$ and $g_\text{sw}$. After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without biochemical downregulation. Hence, $A_\text{amphi}$ and $A_\text{hypo}$ were both measured at low chamber humidity after the leaf had acclimated to high humidity. All other environmental conditions in the leaf chamber remained the same. We logged data until $g_\text{sw}$ reached its nadir. We then repeated the process with the other treatment (amphi or pseudohypo). We did not detect an effect of treatment order on $\mathrm{AA}$ (results not shown).

@@

Coastal ʻilima are surrounded by shorter vegetation than their montane counterparts (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^Veg'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^Veg'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^Veg'))[['p']], 3)`$). The montane site with the lowest vegetation height is a remnant dry forest (Koaiʻa tree sanctuary) in a matrix of cattle pasture, hence the satellite derived vegetation height may be lower than what existed prior to human disturbance. Habitats did not differ significantly in average solar radiation at the top of the canopy (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^Sol'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^Sol'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^Sol'))[['p']], 3)`$). Coastal habitats are significantly warmer (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^MAT'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^MAT'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^MAT'))[['p']], 3)`$) but mean annual precipitation is similar (\autoref{fig:study-system}d; Welch Two Sample *t*-test, $t_{`r round(filter(df_stat, str_detect(name, '^MAP'))[['df']], 2)`} = `r round(filter(df_stat, str_detect(name, '^MAP'))[['statistic']], 2)`$, $P = `r round(filter(df_stat, str_detect(name, '^MAP'))[['p']], 3)`$).

@@

```{=tex}
\begin{figure}[H]
  \includegraphics{../figures/habitat-aa.pdf}
  \caption{Coastal leaves benefit more from amphistomy than montane leaves. A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$. Each point-interval is the median posterior estimate plus 95\% confidence interval of amphistomy advantage for that leaf. Each leaf is from a different montane (blue) or coastal (orange) site, arranged by habitat and ascending amphistomy advantage within habitat. The longer horizonal bars are the average amphistomy advantage for montane and coastal leaves. $g_\mathrm{sw}$, stomatal conductance to water vapor.}
  \label{fig:habitat-aa}
\end{figure}
```

## Leaf thickness is associated with amphistomy advantage between but not within habitats ---

@@

Amphistomy is a seemingly simple way that leaves can increase carbon gain without significant additional water loss, yet it is rare in nature and we do not know why. The strong association between amphiostomy and sunny, open habitats suggests that amphistomy may benefit sun leaves more than shade leaves, but progress has been limited by the lack of evidence that amphiostomy *per se* affects improves photosynthesis in a given leaf. By experimentally blocking gas exchange through the upper surface in a controlled environment, we directly compared an amphistomatous leaf to an otherwise identical pseudohypostomatous leaf. This allows us to quantity the amphistomy advantage ($\mathrm{AA}$) holding all else constant. Taking advantage of the steep climatic gradients in the Hawaiian archipelago, we applied this new method to show for the first time that sun leaves benefit  `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy than shade leaves on ʻilima (*Sida fallax*) plants ($\mathrm{AA}_\text{coastal} = `r prettyNum(habitat_aa[['coastal']], digits = 2L)`$ vs. $\mathrm{AA}_\text{montane} = `r prettyNum(habitat_aa[['montane']], digits = 2L)`$). Coastal and montane ʻilima leaves are likely good representatives of classic sun and shade leaf syndromes because 1) they vary in traits like reflective pubescence [@ehleringer_pubescence_1978] and leaf thickness [@terashima_why_2001] that typically characterize sun-shade adaptation; and 2) since ʻilima shrubs are typically $<1$m tall, they are shaded by trees in montane, but not coastal habitats (\autoref{fig:study-system}d). While this result has not yet been validated in other species, our results indicate that part of the reason amphistomatous leaves are found most commonly in high light habitats is that the adaptive benefit is greater in this environment.

If $\mathrm{AA}$ is typically greater in sun leaves than shade leaves, it could partially explain the distribution of amphi- and hypostoamtous leaves, but the precise mechanism(s) require further study. One hypothesis is that the internal airspace conductance, $g_\mathrm{ias}$, from stomata to mesophyll cell walls is lower in thicker sun leaves [@parkhurst_adaptive_1978]. All else being equal, a leaf with lower $g_\mathrm{ias}$ will benefit more from amphistomy. Our results partially support this hypothesis. Coastal ʻilima leaves with high $\mathrm{AA}$ (\autoref{fig:habitat-aa}) are thicker than montane leaves, but the relationship between $\mathrm{AA}$ and leaf thickness within habitats is actually slightly negative (\autoref{fig:traits-aa}a), opposite our prediction. Since coastal and montane leaves differ in many respects besides thickness, we do not have enough data to conclude that leaf thickness explains the variation in $\mathrm{AA}$ between habitats. Alternatively, other biochemical or anatomical differences between coastal and montane leaves may explain why $\mathrm{AA}$ is greater in coastal leaves. The negative relationship, (albeit nonsignificant in that 95\% confidence intervals encompassed 0), between leaf thickness and $\mathrm{AA}$ could be explained if thicker leaves compensated by having a more porous mesophyll and/or less tortuous airspaces [@theroux-rancourt_maximum_2021].

A second natural hypothesis is that amphistomatous leaves with few adaxial (upper) stomata benefit less than those with similar densities on both surfaces. We predicted that leaves with $g_\mathrm{smax,ratio}$ closer to 0.5 would have higher $\mathrm{AA}$ based on biophysical models [@gutschick_photosynthesis_1984]. The logic is that a small number of stomata on the upper surface are insufficient to supply the entire upper mesophyll due to limited lateral diffusion [@morison_lateral_2005]. Our results do not support this hypothesis. Montane leaves from Big Island sites had low $g_\mathrm{smax,ratio}$ and low $\mathrm{AA}$ whereas low montane leaves on Oʻahu had high $g_\mathrm{smax,ratio}$, but similarly low $\mathrm{AA}$ (\autoref{fig:traits-aa}b). Among coastal sites, the site with the lowest $g_\mathrm{smax,ratio}$ had the lowest $\mathrm{AA}$, but there was little variation in $g_\mathrm{smax,ratio}$ among coastal leaves in our sample. We therefore cannot rule out that a larger sample of coastal leaves with greater variance in $g_\mathrm{smax,ratio}$ might support this hypothesis.

@@

Three limitations of this study are the small sample size, experimental design that precludes distinguishing genetic from environmental differences in leaf traits, and potentially confounding effects of other environmental differences besides light environment. Understanding the mechanistic basis of higher $\mathrm{AA}$ in sun leaves would require much larger sample sizes. Sun leaves tend to be thicker, more densely packed with mesophyll cells, and have greater photosynthetic capacity and higher stomatal conductance, among other traits [@lambers_plant_2008]. Each of these factors and others potentially modulate $\mathrm{AA}$. Quantifying the contribution of all these factors requires larger samples and additional measurements that are beyond the scope of this study, but exciting avenues for future research on leaf structure-function relations. Although many morphological traits that distinguish coastal and montane ʻilima populations persist in a common environment [@yorkston_interfertility_2006], we cannot distinguish between genetic effects and plastic responses to habitat as causes of difference in $\mathrm{AA}$ because we measured naturally occurring plants *in situ*. While disentangling genetic and plastic contributions is not necessarily important for understanding the distribution of amphistomatous leaves, it would be insightful to know about genetic and environmental contributions to trait variation. A reciprocal transplant would be able to determine the genetic and environmental contributions, as well their interaction, to trait variance in nature. However, reciprocal transplants cannot control for other differences between coastal and montane habitats besides vegetation height, such as temperature. Experimental studies in controlled environments will be necessary to isolate the effect light quantity and quality on $\mathrm{AA}$.
